---
title: "We need to find a good title guysss"
author: |
  Maya Archer (25 %), Saumnya (25 %),  
  Rishi Ashok Kumar 560527 (25 %), Nicolas Gonzalez 780037 (25%)
date: "September, 2025"
subtitle: FEM11149 - Introduction to Data Science
output: pdf_document
header-includes:
  - \usepackage{booktabs}
  - \usepackage{makecell}
editor_options: 
  chunk_output_type: inline
urlcolor: blue
linkcolor: red
---

# 1. Introduction

Economic growth is a central concern for governments worldwide, especially when planning to increase public spending that may require external borrowing. Since borrowed funds must eventually be repaid, it is crucial to ensure that the economy grows at a sustainable pace. A commonly used indicator for this purpose is the **Gross Domestic Product (GDP) growth rate**, which provides a measure of how rapidly an economy is expanding over time.  In this project, we work as part of a **data science consulting team** for governmental agencies. One of the agencies has tasked us with investigating whether a country’s GDP growth can be predicted using a combination of **economic and demographic variables**. 
 

```{r Libraries and data loading., include=FALSE}
library(rio)
library(tidyverse)
library(ggplot2)
library(car)
options(scipen = 999)
rm(list = ls())

# Importing Data
df <- import("Data/a1_data_group_9.csv")
```

```{r Data Cleaning and sanity checks, include=FALSE}
# Variable Selection and Renaming
df1 <- df %>% 
  select(`Access to electricity (% of population)`, 
         `Age dependency ratio (% of working-age population)`, 
         `Contributing family workers, total (% of total employment) (modeled ILO estimate)`, 
         `Employers, total (% of total employment) (modeled ILO estimate)`, 
         `Employment in industry (% of total employment) (modeled ILO estimate)`, 
         `Employment in services (% of total employment) (modeled ILO estimate)`, 
         `GDP growth (annual %)`, 
         `Individuals using the Internet (% of population)`, 
         `Labor force participation rate, total (% of total population ages 15+) (modeled ILO estimate)`, 
         `Own-account workers, total (% of male employment) (modeled ILO estimate)`,  
         `Population ages 15-64 (% of total)`, 
         `Population growth (annual %)`, 
         `Tax payments (number)`, 
         `Urban population (% of total)`,  
         `Net trade in goods and services (BoP, current US$)`, 
         `Private credit bureau coverage (% of adults)`, 
         `Public credit registry coverage (% of adults)`) %>% 
  rename(
    electricity_access   = `Access to electricity (% of population)`,
    age_dependency_ratio = `Age dependency ratio (% of working-age population)`,
    family_workers_total = `Contributing family workers, total (% of total employment) (modeled ILO estimate)`,
    employers_total      = `Employers, total (% of total employment) (modeled ILO estimate)`,
    gdp_growth           = `GDP growth (annual %)`,
    internet_users       = `Individuals using the Internet (% of population)`,
    lfpr_total           = `Labor force participation rate, total (% of total population ages 15+) (modeled ILO estimate)`,
    own_account_total    = `Own-account workers, total (% of male employment) (modeled ILO estimate)`,
    pop_15_64_pct        = `Population ages 15-64 (% of total)`,
    pop_growth           = `Population growth (annual %)`,
    tax_payments_num     = `Tax payments (number)`,
    urban_population_pct = `Urban population (% of total)`,
    net_trade            = `Net trade in goods and services (BoP, current US$)`,
    credit_coverage_private = `Private credit bureau coverage (% of adults)`,
    credit_coverage_public = `Public credit registry coverage (% of adults)`
  )

str(df1)

colSums(is.na(df1))
```

# Data 

For this purpose, we rely on the **Global Jobs Indicators Database from the World Bank**, which provides extensive country-level data on employment, trade, demographics, and development indicators. The dataset we use (`a1_data_group_x.csv`) contains **150 observations** (countries) and **74 variables**, including the variable of interest: GDP growth.

After some data exploration, we decided to 
```{r Data Information , include= FALSE}

colnames(df)

```

# Results.


```{r Regression with and Without, echo=FALSE}
fit_with <- lm(
  gdp_growth ~ electricity_access + age_dependency_ratio +
    family_workers_total + employers_total + internet_users +
    lfpr_total + own_account_total + pop_15_64_pct +
    pop_growth + tax_payments_num + urban_population_pct +
    net_trade + credit_coverage_private + credit_coverage_public,
  data = df1,
  singular.ok = FALSE
)

# Model without Net Trade
fit_without <- update(fit_with, . ~ . - net_trade)

# Compare models
anova(fit_without, fit_with)

# Check multicollinearity
vif(fit_with)
```

# Methods. 

# Results.

# Conclusion and Discussion. 

# Code. 

```{r, echo = FALSE, eval = FALSE, warning = FALSE, message = FALSE}
# Directory setup
path = dirname(rstudioapi::getSourceEditorContext()$path) # Path is directory of this file
setwd(path)                                               # Set working directory
```

# 3. Having a bibliography file

In case you use references, you can include them in the RMarkdown file in an automatic way. You see on Canvas a file named `references.bib`, and the file is being called on the header of the `*.rmd` document that generated the pdf. If you open the bib file (you can use either a TeX editor or notepad) you will see a single reference, that has a [BibTeX](https://www.economics.utoronto.ca/osborne/latex/BIBTEX.HTM) format. 

```{}
@book{ISL,
  title={An introduction to statistical learning},
  author={James, Gareth and Witten, Daniela and Hastie, Trevor and Tibshirani, Robert},
  volume={112},
  year={2013},
  publisher={Springer}
}
```

The first thing that appear after the bracket is the alias that you use to 'call' the book in your text. In RMarkdown, you need to use the `@` sign, followed by this alias:`[@ISL]` results in [@ISL]. When you call a reference, RMarkdown will automatically add it in the very end of the document. So we recommend to end your file with a title format for references (even if not is being displayed in your code, it will appear after kniting the file).

You can check on top of this file how the header is organized and where the file with references is being called.

To get a citation already in BibTeX format, you can search for the book or paper in Google Scholar, then click on the quotes that appear below the reference, and click in BibTeX on the pop-up. This will go to another page with only the code, that you can copy and paste in your `references.bib` file.

If you do not want to use a bibliography, then just don't call any reference in the file and erase in the header the command calling `references.bib`.


# 3. Conclusions

You might need more sections than three.


# References
